---
title: "Compare RMATS RI outputs between all tissues normal vs cancer"
output:
  html_document:
    df_print: paged
---

This purpose of this script is to compare intron retention/splicing between normal and cancer samples across all tissues analysed as part of our pipeline. 

Things to note:

* as part of the RMATS config, in this pipeline b1=cancer and b2=normal
* IncLevelDifference = average(inclusion_in_b1)-average(inclusion_in_b2)
* therefore IncLevelDifference>0 = more retention in cancer, IncLevelDifference<0 = more splicing in cancer

Load libraries

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
library(tidyr)
library(duckdb)
library(arrow)
library(stringr)
```
Create DuckDB connection

```{r}
duckdb_connection = dbConnect(duckdb())
```

We will analyse data based on "Junction Count" (JC) differences as opposed to "Junction Count + Exon Count" (JCEC)

```{r}
quant_dir  = "../../../all_TCGA/quant/"
selection = "ID, GeneID, geneSymbol, chr, strand, upstreamEE, downstreamES, IncLevelDifference, FDR"
getFromDDB = function(tissue_type){
  dbGetQuery(duckdb_connection, paste0("SELECT ", selection, " FROM read_parquet('", quant_dir, tissue_type, "/parquet/rMATS_ri/*.parquet')")) %>% mutate(tissue=str_to_sentence(tissue_type))
}

bladder = getFromDDB("bladder")
brain = getFromDDB("brain")
breast = getFromDDB("breast")
cervical = getFromDDB("cervical")
colon = getFromDDB("colon")
head_and_neck = getFromDDB("head_and_neck")
kidney = getFromDDB("kidney")
liver = getFromDDB("Liver")
lung = getFromDDB("lung")
pancreas = getFromDDB("pancreas")
prostate = getFromDDB("prostate")
rectal = getFromDDB("rectal")
skin = getFromDDB("skin")
stomach = getFromDDB("stomach")
thyroid = getFromDDB("thyroid")
uterus = getFromDDB("uterus")
```

We want to merge these into a single data table.

There are a lot of instances where the upstream/downstream exons are slightly larger/smaller but the intron retention event (and JC) is the same. Seeing as we are interested in this at a junction level, we can ignore these columns and call distinct to remove duplicates so we only have 1 row per IR event per tissue. 

```{r}
all_tissue = rbind(bladder, brain, breast, cervical, colon, head_and_neck, kidney, liver, lung, pancreas, prostate, rectal, skin, stomach, thyroid, uterus) %>% 
  transmute(chr=chr,
            strand=strand,
            upstreamEE=upstreamEE,
            downstreamES=downstreamES,
            IncLevelDifference=IncLevelDifference,
            FDR=FDR,
            tissue=tissue) %>% distinct()
rm(bladder, brain, breast, cervical, colon, kidney, liver, lung, ovary, prostate, skin, stomach, thyroid, uterus, head_and_neck, pancreas, rectal)
```

We want to mutate to create a column stating whether there is increased IR in cancer

```{r}
all_tissue = all_tissue %>% mutate(more_IR_in_cancer=IncLevelDifference>0)
```

Now lets create a simple table to show true vs false compared between tissues 

```{r}
all_tissue %>% na.omit() %>%group_by(tissue) %>% summarize(retained_more = sum(more_IR_in_cancer==T),
                                              spliced_more = sum(more_IR_in_cancer==F))
```

How does this compare with non_PTC 3UIs 

```{r}
#read in nonPTC bed file and just isolate the intron coords
#set nonPTC to true
nonPTCs = fread("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.no_cds.bed.gz") %>% transmute(chr=V1, strand=V6,upstreamEE=V2, downstreamES=V3, nonPTC=T) %>% distinct()
#join with all tissues, nonPTC will set to T, any which are F will be set to NA
all_tissue = left_join(all_tissue, nonPTCs) %>% distinct()
#set NAs to F
all_tissue[is.na(all_tissue)]=F
```

Summarize again but subset for nonPTCs

```{r}
all_tissue %>% group_by(tissue, nonPTC) %>% summarize(retained_more = sum(more_IR_in_cancer==T),
                                                      spliced_more = sum(more_IR_in_cancer==F))
```

Lets plot this

```{r, fig.width=5, fig.height=3.5}
all_tissue %>% 
  filter(FDR<0.05) %>%
  filter(abs(IncLevelDifference)>0.05) %>%
  group_by(tissue, nonPTC) %>% summarize(retained_more = sum(more_IR_in_cancer==T),
                                                      spliced_more = sum(more_IR_in_cancer==F)) %>%
    pivot_longer(c(retained_more, spliced_more), names_to="direction") %>%
  mutate(direction=ifelse(direction=="retained_more", "\U2191 Retention", "\U2191 Splicing"),
         nonPTC=ifelse(nonPTC==T, "nonPTC", "pPTC"),
         tissue=ifelse(tissue=="Head_and_neck", "Head/Neck", tissue)) %>%
  filter(tissue!="Skin") %>% 
  ggplot(aes(x=factor(nonPTC, levels=c("pPTC", "nonPTC")),  y=value, fill=factor(direction, levels=c("\U2191 Retention", "\U2191 Splicing")))) + geom_col(position="fill") + 
  facet_wrap(~tissue, nrow = 3) + 
  theme_bw() +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(size=8),
        axis.text = element_text(size=8),
        axis.title.x=element_text(size=8),
        strip.text.x = element_text(size=7.5),
        strip.background = element_rect(fill="white"),
        legend.text = element_text(size=8),
        legend.position="top",
        legend.box.margin = margin(-7.5,0,-7.5,0),
        legend.key.height = unit(0.2,"cm")) + 
  scale_fill_manual(values=c("#999999", "orange")) +
  labs(x="Classification",
       y="") + 
  scale_y_continuous(labels=scales::percent) + 
  geom_hline(yintercept = 0.5, lty=2, col="black") 
ggsave("significant_splicing_between_tissues.png", width=5, height=3.5, units="in")
```

Lets also create a plot just for colon for Figure 2 

```{r}
library(eply)
colon = dbGetQuery(duckdb_connection, "SELECT DISTINCT GeneID, geneSymbol, chr, strand, upstreamEE, downstreamES, IncLevelDifference 
                                       FROM read_parquet('../TCGA_GTEx/utrons/all_TCGA/quant/colon/parquet/rMATS_ri/*.parquet')
                                       WHERE FDR < 0.05")
colon$GeneID = unquote(colon$GeneID)
colon$geneSymbol = unquote(colon$geneSymbol)
colon = left_join(colon, class) %>% distinct()

colon_summary = colon %>% mutate(tempName = paste0(chr,",",strand,",",upstreamEE,",",downstreamES)) %>%
    group_by(tempName,IncLevelDifference) %>%
  summarize(GeneID = ifelse(dplyr::first(match_gene_id)=="",dplyr::nth(n=2,match_gene_id),dplyr::first(match_gene_id))) %>% 
  ungroup() %>% transmute(tempName = tempName, GeneID=GeneID, IncLevelDifference=IncLevelDifference)
```

```{r}
library(org.Hs.eg.db)
colon_summary$symbol = mapIds(org.Hs.eg.db,
                              keys=colon_summary$GeneID,
                              keytype = "ENSEMBL",
                              column="SYMBOL",
                              multiVals = "first")
```
Plot this 

```{r, fig.height=3.5, fig.width=2.5}
library(ggrepel)
label_me = c(1, 2, 3, 4, 5, 10, 12, 14, 36, 340,339,335,330,322)
plot_data = colon_summary %>% left_join(nonPTCs) %>% 
  filter(nonPTC) %>%
  arrange(IncLevelDifference) %>%
  mutate(position = row_number(),
         label=position%in%label_me)

plot_data %>% ggplot(aes(x=fct_reorder(tempName, -IncLevelDifference), 
             y=IncLevelDifference, 
             color=abs(IncLevelDifference)>0.05,
             label=ifelse(label==T&abs(IncLevelDifference)>0.05, symbol,""))) +
  geom_point(size=0.5) + 
  coord_flip() +
  theme_bw(base_size=8) + 
  theme(axis.text.y = element_blank(),
        axis.text.x = element_text(size=8),
        axis.title = element_text(size=8),
        panel.grid.major.y = element_blank(),
        legend.position = "none",
        axis.ticks.y=element_blank()) +
  xlab("Differential 3UI splicing events") + 
  scale_color_manual(values=c("#999999", "orange")) + ylim(-0.5, 0.5) + 
  geom_text_repel(size=2,col="black", arrow=arrow(length=unit(0.01, "npc")),max.overlaps =Inf, segment.size=0.3, segment.color="grey50", box.padding = 0.5, fontface="bold", alpha=1, min.segment.length = unit(0, 'lines'))
ggsave("over_vs_undersplicing_coad.png", width=2.5, height=3.5, units = "in")
```