---
title: "Analyse rmats output"
output:
  html_document:
    df_print: paged
---

Analyse RMATS outputs

Load packages

```{r}
library(stringr)
library(dplyr)
library(data.table)
library(tidyr)
library(ggplot2)
```

IncLevelDif= a - b 

a = 0uM
b = 20uM

If 50% spliced in a, and 100% spliced in b, then incleveldif = 0.5 - 0 = 0.5. Therefore >0 = spliced more, <0 = ret more. 

#Setup

```{r}
output = fread("../../../../../HCT_CHIR/RMATS/outputs/SIG_RI.MATS.JC.txt")
output = subset(output, select=which(!duplicated(names(output)))) %>%  transmute(chr=chr,
            strand=strand,
            upstreamEE=upstreamEE,
            downstreamES=downstreamES,
            IncLevelDifference=IncLevelDifference, 
            PValue) %>% distinct() %>%
  mutate(FDR=p.adjust(PValue, method="BH"))
```


```{r}
table(spliced_more=output[,5]>0)
```
This contains PTC and non-PTC 3UIs. How does it compare if we subset based on this

```{r}
nonPTCs = fread("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.no_cds.bed.gz")
```

Also remove dupes from this so it runs quicker, we only need the 2 coors so get rid of other cols

```{r}
nonPTCs = nonPTCs %>% transmute(V2=V2, V3=V3) %>% distinct() %>% mutate(nonPTC=T)
```

Now join together and set NA to FALSE 

```{r}
unique_nonPTC = left_join(output, nonPTCs, by=c("upstreamEE"="V2", "downstreamES"="V3"))
unique_nonPTC[is.na(unique_nonPTC)]=F
```

### IncLevel vs IncLevel (like the DTU plot from Fig4)

```{r, warning=F, message=F}
library(stringr)
rmats_output = fread("../../../../../HCT_CHIR/RMATS/outputs/RI.MATS.JC.txt") %>% select(chr=chr,
            strand=strand,
            upstreamEE=upstreamEE,
            downstreamES=downstreamES,
            IncLevel1=IncLevel1,
            IncLevel2=IncLevel2,
            IncLevelDifference) %>% distinct()

list_to_mean = function(list1){
  unlisted1 = str_split(list1, ",")[[1]] 
  unlisted1 = unlisted1[!is.na(unlisted1)] %>% 
    as.numeric() %>% na.omit() %>%
    mean()
  return(unlisted1)
}
rmats_output = rmats_output %>% rowwise() %>% mutate(mean_1 = list_to_mean(IncLevel1),
                                      mean_2 = list_to_mean(IncLevel2))
```

```{r, fig.height=3.5, fig.width=3.5}
output = fread("../../../../../HCT_CHIR/RMATS/outputs/sig_RI.MATS.JC.txt") %>% select(chr=chr,
            strand=strand,
            upstreamEE=upstreamEE,
            downstreamES=downstreamES) %>% distinct() %>% mutate(sig=T)
rmats_output = rmats_output %>% left_join(output)
rmats_output = rmats_output %>% mutate(sig=ifelse(is.na(sig), F, T))
```

Specifically plot the Wnt components

```{r, fig.height=2.5, fig.width=2.5}
canonical_wnt_genes = fread("../../../../../HCT_CHIR/RMATS/canonical_wnt_genes.csv")
tx2gene = fread("../../../all_TCGA/expression.dir/csvdb_files/tx2gene.txt")
wnt_tx = tx2gene %>% left_join(canonical_wnt_genes, by=c("match_gene_id"="GeneID")) %>% na.omit()
bed = fread("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.all_utrons.bed.gz")
wnt_bed = bed %>% left_join(wnt_tx, by=c("V4"="transcript_id")) %>% na.omit()

filter_frame = wnt_bed %>% select(upstreamEE=V2,
                                  downstreamES=V3,
                                  wnt_tx_id=V4,
                                  symbol=symbol) %>% distinct()

library(ggrepel)
rmats_output %>% 
  left_join(filter_frame) %>% 
  filter(!is.na(wnt_tx_id)) %>% select(-wnt_tx_id) %>% distinct() %>%
  ggplot(aes(x=mean_1, y=mean_2, color=sig, alpha=ifelse(sig, 1, 0.5), label=ifelse(sig, symbol,""))) + geom_point() + scale_color_manual(values=c("black", "red")) + theme_bw(base_size=8) + 
  theme(legend.position = "none",
        axis.text = element_text(size=8)) + 
  labs(x="IncLevel in 0uM",
       y="IncLevel in 20uM") + geom_abline(intercept = 0, lty=2) + geom_text_repel(size=2.5,col="black", fontface="bold", arrow=arrow(length=unit(0.01, "npc")),max.overlaps =21, segment.size=0.3, segment.color="grey50")+ 
  annotate("text", x = 0.2,  y=0.875, label="Retained more", fontface="bold", size=2.5) + 
  annotate("text", x = 0.775,  y=0.175, label="Spliced more", fontface="bold", size=2.5)+
  annotate("text", x = 0.2,  y=0.825, label="with \u2191 Wnt", fontface="bold", size=2.5) + 
  annotate("text", x = 0.775,  y=0.125, label="with \u2191 Wnt", fontface="bold", size=2.5)

ggsave("wnt_components_rmats_dotplot.png", height=2.5, width=2.5, units="in")
```

### Lollipop chart for incLevelDifference (ordered most + to most -)

```{r}
library(duckdb)
library(arrow)
duckdb_connection = dbConnect(duckdb())
quant_dir  = "../../../all_TCGA/quant/"

wnt = unique_nonPTC %>% filter(nonPTC==T, IncLevelDifference>0) %>% mutate(tempName=paste0(upstreamEE,",",downstreamES))

selection = "chr, strand, upstreamEE, downstreamES, IncLevelDifference, IncLevel1, IncLevel2, PValue"
getFromDDB = function(tissue_type){
  dbGetQuery(duckdb_connection, paste0("SELECT ", selection, " FROM read_parquet('", quant_dir, tissue_type, "/parquet/rMATS_ri/*.parquet')")) %>% mutate(tissue=str_to_sentence(tissue_type))
}

colon_wnt = getFromDDB("colon") %>% 
  distinct() %>% 
  mutate(tempName=paste0(upstreamEE, ",", downstreamES)) %>% 
  filter(tempName %in% wnt$tempName) %>% 
  rowwise() %>% 
  mutate(mean_1 = list_to_mean(IncLevel1),
        mean_2 = list_to_mean(IncLevel2),
        padj = p.adjust(PValue, method="BH"))
```

```{r, fig.height=2.5, fig.width=2.5}
library(forcats)

plotdata = getFromDDB("colon") %>% distinct() %>% mutate(tempName=paste0(upstreamEE,",",downstreamES)) %>%
  filter(tempName %in% wnt$tempName) %>% distinct() 

more_spliced = plotdata %>% filter(IncLevelDifference<0)
more_ret = plotdata %>% filter(IncLevelDifference>0)

colon_wnt %>% arrange(-IncLevelDifference) %>% ggplot(aes(x=fct_reorder(tempName, -IncLevelDifference), y=IncLevelDifference, color=IncLevelDifference>0)) + 
  geom_segment(aes(x=fct_reorder(tempName, -IncLevelDifference), xend=fct_reorder(tempName, -IncLevelDifference), y=0, yend=IncLevelDifference)) + 
  theme_bw(base_size=8) + 
  theme(axis.text.x = element_text(color="white"),
        axis.ticks.x = element_blank(),
        panel.grid.major = element_blank(),
        axis.text = element_text(size=8),
        legend.position = "none") + 
  labs(x="3UIs overspliced upon Wnt activation",
       y="IncLevelDifference in Cancer vs Healthy") + 
  annotate("text", x = 458,  y=-0.075, hjust=1, label="Spliced more", fontface="bold", size=2.5) + 
  annotate("text", x = 50,  y=0.125, hjust=0,label="Retained more", fontface="bold", size=2.5)+
  annotate("text", x = 458,  y=-0.1, hjust=1,label="in Cancer", fontface="bold", size=2.5) + 
  annotate("text", x = 50,  y=0.1, hjust=0,label="in Cancer", fontface="bold", size=2.5) + 
  annotate("text", x = 50,  y=0.075, hjust=0,label=paste0("(",round(nrow(more_ret)*100/nrow(plotdata), 1), "%)"), fontface="bold", size=2.5) +
  annotate("text", x = 458,  y=-0.125, hjust=1,label=paste0("(",round(nrow(more_spliced)*100/nrow(plotdata), 1), "%)"), fontface="bold", size=2.5) + 
  scale_color_manual(values=c("orange", "#999999")) + 
  geom_vline(xintercept=nrow(colon_wnt)/2, lty=2, color="gray50", alpha=0.5)

ggsave("lollipop_chart_colon_wnt.png", height=2.5, width=2.5, units="in")
```


### Plot colon-vs-healthy vs wnt-manipulation by signature

```{r}
getFromDDB = function(tissue_type){
  dbGetQuery(duckdb_connection, paste0("SELECT ", selection, " FROM read_parquet('", quant_dir, tissue_type, "/parquet/rMATS_ri/*.parquet')")) %>% mutate(tissue=str_to_sentence(tissue_type))
}
colon = getFromDDB("colon") %>% distinct() %>% mutate(FDR=p.adjust(PValue, method="BH")) %>% filter(FDR<0.05) %>% left_join(nonPTCs, by=c("upstreamEE"="V2", "downstreamES"="V3"))
colon[is.na(colon)]=F
colon_summary = colon %>% mutate(more_IR_in_canc = IncLevelDifference>0) %>% group_by(nonPTC) %>% 
  summarize(retained_more = sum(more_IR_in_canc==T),
            spliced_more = sum(more_IR_in_canc==F))
colon_summary = colon_summary %>% pivot_longer(c(retained_more, spliced_more), names_to="direction")
colon_summary

colon = getFromDDB("colon") %>% distinct() %>% left_join(nonPTCs, by=c("upstreamEE"="V2", "downstreamES"="V3"))
colon[is.na(colon)]=F

colon_summary = colon %>% mutate(more_IR_in_canc = IncLevelDifference>0) %>% group_by(nonPTC) %>% 
  summarize(retained_more = sum(more_IR_in_canc==T),
            spliced_more = sum(more_IR_in_canc==F))
colon_summary = colon_summary %>% pivot_longer(c(retained_more, spliced_more), names_to="direction")
colon_summary

output = fread("../../../../../HCT_CHIR/RMATS/outputs/RI.MATS.JC.txt")
output = subset(output, select=which(!duplicated(names(output)))) %>%  transmute(chr=chr,
            strand=strand,
            upstreamEE=upstreamEE,
            downstreamES=downstreamES,
            IncLevelDifference=IncLevelDifference) %>% distinct() 

unique_nonPTC = left_join(output, nonPTCs, by=c("upstreamEE"="V2", "downstreamES"="V3"))
unique_nonPTC[is.na(unique_nonPTC)]=F

final = unique_nonPTC %>% mutate(more_IR_in_treated = IncLevelDifference<0) %>% group_by(nonPTC) %>% 
  summarize(retained_more = sum(more_IR_in_treated==T),
            spliced_more = sum(more_IR_in_treated==F)) %>%
  pivot_longer(c(retained_more, spliced_more), names_to="direction")
```

```{r, fig.height=2.5, fig.width=2.5}
colon_summary = colon_summary %>% mutate(tissue="Colon")
final = final %>% mutate(tissue="Wnt activation")
final = rbind(final, colon_summary)

final %>% mutate(direction = ifelse(direction=="spliced_more", "Spliced more", "Retained more"),
                 nonPTC=ifelse(nonPTC==T, "NonPTC", "pPTC")) %>% 
  ggplot(aes(x=factor(nonPTC, levels=c("pPTC", "NonPTC")),  y=value, fill=factor(direction, levels=c("Retained more", "Spliced more")))) + geom_col(position="fill", col="black") + 
  theme_bw(base_size = 8) +
  theme(aspect.ratio = 1.5, 
        legend.title = element_blank(),
        axis.text.x = element_text(size=8, angle=0, hjust=0.5),
        strip.text.x = element_text(size=8),
        strip.background = element_rect(fill="white"), 
        legend.position = "top",
        legend.text = element_text(size=8)) + 
  scale_fill_manual(values=c("#999999", "orange")) +
  labs(x="Classification",
       y="") + 
  scale_y_continuous(labels=scales::percent) + 
  facet_wrap(~tissue)

ggsave("colon_vs_wnt_signature.png", height=2.5, width=2.5, units="in")
```