---
title: "Supplementary Figure 6B - UPF1 knockdown effect on HRAS reporters"
output:
  html_document:
    df_print: paged
---

## Background

HCT116 cells were seeded Thurs 20th Oct 2022 into 24 well plates at 35,000 cells / well

RNAi done on Saturday 22 Oct, They were ~30% confluent. pDNA transfection done 23 Oct (70% confluent). 2nd hit RNAi done 24 Oct. 


Luciferase was measured using luminometer in SW lab. Followed standard protocol but used reduced amount of reagents:
 - passive lysis in 100ul PLB
 - 25ul LAR II (instead of 100ul)
 - 5ul of lysate (instead of 20ul)
 - 25ul Stop & Glo (instead of 100ul)
 
## Code

```{r load, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(data.table)
library(tidyr)
library(ggpubr)

data = fread("raw_data_UPF1_Luc.csv")
```



```{r}
Luc_data = data %>% filter(Luc_type=="Firefly") %>% select(-Luc_type)
Ren_data = data %>% filter(Luc_type=="Renilla") %>% select(-Luc_type)
names(Luc_data)[4]="Luc_value"
names(Ren_data)[4]="Ren_value"

joined_data = left_join(Luc_data, Ren_data) %>% 
  mutate(Ratio=Luc_value/Ren_value) %>%
  group_by(RNAi,Isoform, Biorep) %>% 
  summarize(mean_Ratio = mean(Ratio))

merged = joined_data %>% group_by(RNAi, Isoform) %>% 
  summarize(sd=sd(mean_Ratio), num=n(), sem=sd/sqrt(num), mean=mean(mean_Ratio))
```

```{r}


cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

```




```{r, fig.height=2.5, fig.width=6}
my_comparisons = list(c("siDsRed", "siUPF1"))
joined_log = joined_data %>% mutate(log_mean = log2(mean_Ratio))
joined_log %>%
  mutate(Isoform = factor(Isoform, levels=c("flUTR", "GT>GA", "deltaIntron")),
         RNAi = sub("siUPF1_2", "siUPF1", RNAi)) %>%
ggbarplot( x="RNAi", 
          y="log_mean", 
          add="mean_se", 
          order=c("siDsRed", "siUPF1"),
          facet.by = "Isoform",
          fill="Isoform",
                    palette = cbPalette) +
  labs(x="RNAi", y="Log2 (Firefly/Renilla)") + 
  theme_bw(base_size=8) + 
  theme(legend.position = "none", axis.text.x=element_text(angle=45, hjust=1)) + scale_fill_manual(values=cbPalette) + 
  stat_compare_means(comparisons=my_comparisons, method="t.test") +
  coord_cartesian(ylim=c(0,4.5)) -> saveme
 png("UPF1_HRAS_Luc.png", width=5, height=2, units="in", res=1200)
 saveme
 dev.off()
```
