---
title: "siUPF1 DTE ECDF plots"
output:
  html_document:
    df_print: paged
---

The purpose of this script is to look at how population classes of 3'UTR spliced transcripts react to UPF1 knockdown

```{r}
library(DESeq2)
library(tximport)
library(dplyr)
library(tibble)
library(stringr)
```

Load sample names

```{r}
tx2gene<-read.delim(file='../../../merged_gtfs/all_tissues_merged/final_geneset.dir/all_tissues_agg-agg-agg.filtered.tx2gene.tsv', 
                    header = TRUE, sep = '\t', dec = '.') %>%
  dplyr::select(transcript_id, gene_id=match_gene_id)

quant_dir = "../../../../../HCT-NMD_KD/requant/quantification.dir"
all_files = list.files(quant_dir, pattern="*")
all_files = file.path(quant_dir,  all_files, "quant.sf")

sample_names <- sapply(all_files, function(x) substr(x, 0, nchar(x)-3))
sample_names <- sapply(sample_names, function(x) substr(x, 32, nchar(x)))
names(all_files) <- sample_names
```

Generate sample table, comparing siUPF1 to control

```{r}
sample_table = data.frame(files=all_files, names=sample_names)
sample_table = sample_table %>% mutate(condition=ifelse(stringr::str_detect(names, "HCT116-R"), "siControl", "siUPF1"))
sample_table$files = as.character(sample_table$files)
```

Tximport transcript level

```{r}
txi <- tximport(sample_table$files, type="salmon", tx2gene=tx2gene, txOut=T)
```

Generate deseq dataset to compare the conditions

```{r}
dds <- DESeqDataSetFromTximport(txi, sample_table, ~condition)
```

Run deseq

```{r}
dds_result = DESeq(dds, parallel = T)
```

Filter for rows where there is at least 1 count in all samples for at least 1 condition

```{r}
nrow(dds_result)
keep = rowSums(counts(dds_result)[,1:8]>1)>=4|rowSums(counts(dds_result)[,9:16]>1)>=4
sum(keep)
dds_result = dds_result[keep,]
nrow(dds_result)
```

Deseq2 result

```{r}
dds_result = results(dds_result)
```

MA plot

```{r}
plotMA(dds_result)
```
Add annotation for 3UIs
```{r}
all_utrons <- read.delim("../../../../final_classifications/utron_beds.dir/TCGA-agg-agg.all_utrons.ids.gz", header=F)
novel_utrons <- read.delim("../../../../final_classifications/utron_beds.dir/TCGA-agg-agg.novel_utrons.ids.gz", header=F)
partnered_utrons <- read.delim("../../../../final_classifications/utron_beds.dir/TCGA-agg-agg.partnered_utrons.ids.gz", header=F)
non_cds <- read.delim("../../../../final_classifications/utron_beds.dir/TCGA-agg-agg.no_cds_utrons.ids.gz", header=F)

utron_annotation <- data.frame(  transcript_id = rownames(dds_result)) %>%
  mutate(has_3UI = transcript_id %in% all_utrons$V1,
         has_novel_3UI = transcript_id %in% novel_utrons$V1,
         has_e3UI = transcript_id %in% non_cds$V1)

suplementary_data <- dds_result %>%
  as.data.frame() %>%
  rownames_to_column("transcript_id") %>%
  left_join(tx2gene) %>%
  left_join(utron_annotation) %>%
  relocate(gene_id)
write.table(suplementary_data, "Supplementary_data_9a.tsv", sep="\t", row.names=F, quote=F)

```



Read in RMATs results

```{r}
library(tidyverse)
read.delim("../../../../../HCT-NMD_KD/requant/rmats.dir/HCT116/RI.MATS.JC.txt") %>%
  dplyr::select(-ID, -ID.1) %>%
  distinct() %>%
  write.table("Supplmentary_data_9c.tsv", sep="\t", row.names=F, quote=F)

```

DO the same for the CHIR seq

```{r}
read.delim("../../../../../HCT_CHIR/RMATS/outputs/RI.MATS.JC.txt") %>%
  dplyr::select(-ID, -ID.1) %>%
  distinct() %>%
  write.table("Supplmentary_data_9d.tsv", sep="\t", row.names=F, quote=F)

```