---
title: "Using mean PSI instead of utroniness"
output: html_notebook
---

The Utroniness measure suffers from the problem that it is not independent of the
total expression in a sample. One way to solve that might be to use the PSI measurements
instead, as these are indeivdually normalised to gene expression, and I think that
they are also shruken. Its unclear what summary stateistic to use, but lets try starting with mean. 


```{r}
library(tidyverse)
all_utrons <- read.delim("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.all_utrons.ids.gz", header=F,
                         col.names = "transcript_id")
all_utrons$non_cds <- all_utrons$transcript_id %in% read.delim("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.no_cds.ids.gz",
                                                               header=FALSE,
                                                               col.names=c("transcript_id", "match_id"))$transcript_id
all_utrons$novel <- all_utrons$transcript_id %in% read.delim("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.novel_utrons.ids.gz",
                                                               header=FALSE,
                                                               col.names=c("transcript_id", "match_id"))$transcript_id
all_utrons$match_id <- all_utrons$transcript_id %in% read.delim("../../../all_TCGA/new_classification/utron_beds.dir/agg-agg-agg.no_cds.ids.gz",
                                                               header=FALSE,
                                                               col.names=c("transcript_id", "match_id"))$match_id
```

```{r}
library(duckdb)
db <- dbConnect(duckdb())
duckdb_register(db, "utrons", all_utrons)
utroniness <- dbGetQuery(db,
                         "SELECT distinct_pso.track, non_cds, condition,  AVG(IFNULL(1-PSI,0)) as utroniness
                         FROM
                         (SELECT DISTINCT track, non_cds, condition, start, \"end\", strand, psi
                         FROM read_parquet('../../../all_TCGA/quant/colon/parquet/rmats_junction_counts/*.parquet') as pso
                           LEFT JOIN utrons ON utrons.transcript_id = pso.transcript_id
                         WHERE  (spliced_reads + retained_reads) > 0 ) as distinct_pso
                         GROUP BY
                         track, non_cds, condition
                         ")
dbDisconnect(db, shutdown=TRUE)
utroniness %>%
  mutate(normal=ifelse(grepl("test", condition), "Cancer", "Normal")) %>%
  ggplot() + aes(utroniness) +
  facet_grid(normal~non_cds) +
  geom_density()
```



```{r}

db <- dbConnect(duckdb())
gene_tpms <- dbGetQuery(db,
                         "SELECT DISTINCT
                            sample_id, 
                            gene_id,
                            gene_tpm
                         FROM 
                           read_parquet('../../../all_TCGA/quant/colon/parquet/utrons_expression/*.parquet')",
                         )
dbDisconnect(db, shutdown=TRUE)

total_tpm <- gene_tpms %>%
  mutate(sample_id = gsub(".", "-", sample_id, fixed=T)) %>%
  group_by(sample_id) %>%
  summarise(total_tpm=sum(gene_tpm, na.rm=T))
```

```{r}
utroniness %>%
  inner_join(total_tpm, by=c("track"="sample_id")) %>% 
  ggplot() + aes(y=utroniness, x=total_tpm) +
  geom_point() +
  stat_smooth(method="lm") +
  facet_grid(non_cds ~ condition) +
  theme_bw()

```


Lets start by looking at the correlations. Starting with UPF1.


```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)
tx2gene <- distinct(AnnotationDbi::select(org.Hs.eg.db, gene_tpms$gene_id, keytype = "ENSEMBL", columns = "SYMBOL"))
gene_tpms %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  inner_join(filter(tx2gene, SYMBOL=="UPF1"), by=c("gene_id"="ENSEMBL")) %>%
  inner_join(utroniness, by=c("sample_id"="track"), relationship="many-to-many") %>%
  group_by(gene_id, non_cds, condition) %>%
  do(with(., cor.test(utroniness, gene_tpm, method="spearman", exact=F)) %>% broom::tidy()) 
```

```{r}
gene_tpms %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  inner_join(utroniness, by=c("sample_id"="track"), relationship="many-to-many") %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>%
  do(with(., cor.test(utroniness, gene_tpm, method="spearman", exact=F)) %>% broom::tidy()) %>%
  left_join(tx2gene, by=c("gene_id"="ENSEMBL")) -> correlations

correlations %>%
  mutate(non_cds=ifelse(non_cds, "nonPTC", "pPTC"),
         condition=ifelse(condition == "test", "Cancer", "Normal"),
         FDR=ifelse(p.adjust(p.value, method="BH") < 0.05, "<0.05", "NS")) %>%
  ggplot() + aes(estimate, fill=FDR) +
  geom_histogram() +
  facet_grid(condition ~ non_cds) +
  geom_vline(xintercept = 0, lty=2, col="grey25") +
  xlab(expression(paste("Spearman's ", rho))) +
  theme_bw()
```

```{r}
correlations %>%
  filter(non_cds, condition=="test") %>%
  dplyr::select(SYMBOL, condition, estimate, p.value) %>%
  arrange(desc(estimate)) %>% head(n=100)

```

```{r}
correlations %>%
  filter(grepl("UPF", SYMBOL) | grepl("SMG", SYMBOL)) %>%
  dplyr::select(SYMBOL, non_cds, condition, estimate, p.value)

```


The general negative correlations between genes and utroniness is puzzling. But one possibility is that
there is some highly expressed gene that is correlated with utroniness, that is absorbing
counts, and thus depressing the counts of every thing else. We could try using DESeq normalisation to deal
with this. 

In DESeq normalisation, we calculate a pseudo reference, by taking the gemetric mean
of expression for each gene. 

```{r}
normed_tpm <- 
  gene_tpms %>%
  group_by(gene_id) %>%
  summarize(geoMean=exp(mean(log(gene_tpm)))) %>%
  right_join(gene_tpms) %>% 
  group_by(sample_id) %>%
  summarize(normFactor = median(gene_tpm/geoMean, na.rm=T)) %>% 
  left_join(gene_tpms) %>%
  mutate(normed_tpm=gene_tpm/normFactor)

```

```{r}
normed_tpm %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  inner_join(utroniness, by=c("sample_id"="track"), relationship="many-to-many") %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>%
  do(with(., cor.test(utroniness, normed_tpm, method="spearman", exact=F)) %>% broom::tidy()) %>%
  left_join(tx2gene, by=c("gene_id"="ENSEMBL")) -> norm_correlations

norm_correlations %>%
  mutate(non_cds=ifelse(non_cds, "nonPTC", "pPTC"),
         condition=ifelse(condition == "test", "Cancer", "Normal"),
         FDR=ifelse(p.adjust(p.value, method="BH") < 0.05, "<0.05", "NS")) %>%
  ggplot() + aes(estimate, fill=FDR) +
  geom_histogram() +
  facet_grid(condition ~ non_cds) +
  geom_vline(xintercept = 0, lty=2, col="grey25") +
  xlab(expression(paste("Spearman's ", rho))) +
  theme_bw()

```

This is much closer to the expected distributin of correlation coefficients. What does this look like for the 
NMD factors?

```{r}
norm_correlations %>%
  mutate(FDR=p.adjust(p.value, method="BH")) %>%
  filter(grepl("^UPF", SYMBOL) | grepl("^SMG", SYMBOL)) %>%
  dplyr::select(SYMBOL, non_cds, condition, estimate, FDR)

```
```{r, fig.height=3, fig.width=3}
norm_correlations %>%
  ungroup() %>%
  mutate(FDR=p.adjust(p.value, method="BH"),
         non_cds=ifelse(non_cds, "nPTC", "pPTC"),
         condition=ifelse(condition=="test", "Cancer", "Normal"),
         category=paste(condition, non_cds)) %>%
  filter(grepl("^UPF", SYMBOL) | grepl("^SMG", SYMBOL)) %>%
  dplyr::select(SYMBOL, category, estimate) %>%
  pivot_wider(values_from=estimate, id_cols=SYMBOL, names_from=category) %>%
  as.data.frame() -> nmd_matrix

rownames(nmd_matrix) <- nmd_matrix$SYMBOL
nmd_matrix <- as.matrix(nmd_matrix[,-1])

library(ComplexHeatmap)
Heatmap(nmd_matrix,
        col=circlize::colorRamp2(c(-0.5,0,0.5), c("Blue", "White", "Red")),
        row_names_gp = grid::gpar(fontsize=8),
        column_names_gp = grid::gpar(fontsize=8),
        column_names_rot=45,
        name=" ",
) -> saveme
png("NMD_components_PSI_correlations.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```

```{r, fig.height=3, fig.width=3}
x = normed_tpm %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  filter(gene_id=="ENSG00000005007") %>% #filter for UPF1
  inner_join(utroniness, by=c("sample_id"="track")) %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>% 
  mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"))

stats_data = norm_correlations %>% 
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000005007") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",\npadj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=40, 
            y=ifelse(non_cds=="nPTC", 0.325, 0.675))
x$condition = factor(x$condition, levels=c("Normal", "Cancer"))
x$non_cds = factor(x$non_cds, levels=c("pPTC", "nPTC"))

x %>%  
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.75) + 
  geom_smooth(method="lm", color="black") +
  facet_grid(vars(factor(non_cds, levels=c("pPTC", "nPTC") )), vars(factor(condition, levels=c("Normal", "Cancer"))), scales="free_y") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("UPF1 Expression") +
  ylab("Average 3UI PSI per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0.5, vjust=0.5, nudge_y = 0.005) -> saveme
png("UPF1_comparison_PSI_correlations.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```

```{r}
#save.image("PSI_correlations_temp.RData")
load("PSI_correlations_temp.RData")
```


### Plots for paper

```{r, fig.height=3, fig.width=3}
norm_correlations %>%
  ungroup() %>%
  mutate(FDR=p.adjust(p.value, method="BH"),
         non_cds=ifelse(non_cds, "nPTC", "pPTC"),
         condition=ifelse(condition=="test", "Cancer", "Normal"),
         category=paste(condition, non_cds)) %>%
  filter(grepl("^UPF", SYMBOL) | grepl("^SMG", SYMBOL)) %>%
  dplyr::select(SYMBOL, category, estimate) %>%
  pivot_wider(values_from=estimate, id_cols=SYMBOL, names_from=category) %>%
  as.data.frame() -> nmd_matrix

rownames(nmd_matrix) <- nmd_matrix$SYMBOL
nmd_matrix <- as.matrix(nmd_matrix[,-1])

library(ComplexHeatmap)
Heatmap(nmd_matrix,
        col=circlize::colorRamp2(c(-0.5,0,0.5), c("Blue", "White", "Red")),
        row_names_gp = grid::gpar(fontsize=8),
        column_names_gp = grid::gpar(fontsize=8),
        column_names_rot=45,
        name=" ",
) -> saveme
png("NMD_components_PSI_correlations.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```

### UPF1 

nPTC and pPTC

```{r, fig.height=3, fig.width=3}
x = normed_tpm %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  filter(gene_id=="ENSG00000005007") %>% #filter for UPF1
  inner_join(utroniness, by=c("sample_id"="track")) %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>% 
  mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"))

stats_data = norm_correlations %>% 
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000005007") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",\npadj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=ifelse(condition=="Normal",30,40), 
            y=ifelse(non_cds=="nPTC", 0.325, 0.675))
x$condition = factor(x$condition, levels=c("Normal", "Cancer"))
x$non_cds = factor(x$non_cds, levels=c("pPTC", "nPTC"))

x %>%  
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.75) + 
  geom_smooth(method="lm", color="black") +
  facet_grid(vars(factor(non_cds, levels=c("pPTC", "nPTC") )), vars(factor(condition, levels=c("Normal", "Cancer"))), scales="free") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("UPF1 Expression") +
  ylab("Average 3UI PSO per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0.5, vjust=0.5, nudge_y = 0.005) -> saveme
png("UPF1_comparison_PSI_correlations_both.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```

nPTC only

```{r, fig.height=2, fig.width=2}
stats_data = norm_correlations %>% 
  filter(non_cds) %>%
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000005007") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",padj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "UPF1 - Normal", "UPF1 - Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=ifelse(condition=="UPF1 - Cancer", 0, 10), 
            y=ifelse(condition=="UPF1 - Cancer", 0.325, 0.4125))

x %>%  
  filter(non_cds=="nPTC") %>%
  mutate(condition=ifelse(condition=="Normal", "UPF1 - Normal", "UPF1 - Cancer")) %>%
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.5) + 
  geom_smooth(method="lm", color="black") +
  facet_wrap(vars(factor(condition, levels=c("UPF1 - Normal", "UPF1 - Cancer"))), nrow=2, scales = "free") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.text.y=element_text(size=7),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("UPF1 Expression") +
  ylab("Average 3UI PSO per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0, vjust=0.5, nudge_y = 0.005) -> saveme
png("UPF1_comparison_PSI_correlations_nPTC_only.png", height=2, width=2, units="in", res=1200)
saveme
dev.off()
```
### UPF3B

ENSG00000125351

nPTC and pPTC

```{r, fig.height=3, fig.width=3}
x = normed_tpm %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  filter(gene_id=="ENSG00000125351") %>% #filter for UPF1
  inner_join(utroniness, by=c("sample_id"="track")) %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>% 
  mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"))

stats_data = norm_correlations %>% 
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000125351") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",\npadj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=ifelse(condition=="Normal",6,15), 
            y=ifelse(non_cds=="nPTC", 0.325, 0.675))
x$condition = factor(x$condition, levels=c("Normal", "Cancer"))
x$non_cds = factor(x$non_cds, levels=c("pPTC", "nPTC"))

x %>%  
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.75) + 
  geom_smooth(method="lm", color="black") +
  facet_grid(vars(factor(non_cds, levels=c("pPTC", "nPTC") )), vars(factor(condition, levels=c("Normal", "Cancer"))), scales="free") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("UPF3B Expression") +
  ylab("Average 3UI PSO per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0.5, vjust=0.5, nudge_y = 0.005) -> saveme
png("UPF3B_comparison_PSI_correlations_both.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```

nPTC only

```{r, fig.height=2, fig.width=2}
stats_data = norm_correlations %>% 
  filter(non_cds) %>%
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000125351") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",padj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "UPF3B - Normal", "UPF3B - Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=17, 
            y=0.325)

x %>%  
  filter(non_cds=="nPTC") %>%
  mutate(condition=ifelse(condition=="Normal", "UPF3B - Normal", "UPF3B - Cancer")) %>%
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.75) + 
  geom_smooth(method="lm", color="black") +
  facet_wrap(vars(factor(condition, levels=c("UPF3B - Normal", "UPF3B - Cancer"))), nrow=2, scales="free") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("UPF3B Expression") +
  ylab("Average 3UI PSI per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0.5, vjust=0.5, nudge_y = 0.005) -> saveme
png("UPF3B_comparison_PSI_correlations_nPTC_only.png", height=2, width=2, units="in", res=1200)
saveme
dev.off()
```
### SMG8

ENSG00000167447

nPTC and pPTC

```{r, fig.height=3, fig.width=3}
x = normed_tpm %>%
  mutate(sample_id=gsub(".", "-", sample_id, fixed=T)) %>%
  filter(gene_id=="ENSG00000167447") %>% #filter for UPF1
  inner_join(utroniness, by=c("sample_id"="track")) %>%
  group_by(gene_id, non_cds, condition) %>%
  filter(sd(utroniness) > 0, sd(gene_tpm) > 0) %>% 
  mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"))

stats_data = norm_correlations %>% 
  mutate(padj = p.adjust(p.value, method="BH")) %>%
  filter(gene_id == "ENSG00000167447") %>% 
  mutate(label = paste0("R=", round(estimate,2), ",\npadj=", formatC(padj, format="e", digits=2))) %>%
           mutate(condition=ifelse(condition=="normal", "Normal", "Cancer"),
            non_cds=ifelse(non_cds==T, "nPTC", "pPTC"),
            x=ifelse(condition=="Normal",4.5,10), 
            y=ifelse(non_cds=="nPTC", 0.325, 0.675))
x$condition = factor(x$condition, levels=c("Normal", "Cancer"))
x$non_cds = factor(x$non_cds, levels=c("pPTC", "nPTC"))

x %>%  
  ggplot(aes(x=normed_tpm, y=utroniness)) + 
  geom_point(color="#999999", alpha=0.75) + 
  geom_smooth(method="lm", color="black") +
  facet_grid(vars(factor(non_cds, levels=c("pPTC", "nPTC") )), vars(factor(condition, levels=c("Normal", "Cancer"))), scales="free") + 
  theme_bw(base_size=8) + 
  theme(axis.text=element_text(size=8),
        axis.title=element_text(size=8),
        strip.background=element_rect(fill="white"),
        strip.text=element_text(size=8)) + 
  xlab("SMG8 Expression") +
  ylab("Average 3UI PSO per sample") + 
  geom_text(data = stats_data, mapping = aes(label=label, x=x, y=y), size=3, hjust=0.5, vjust=0.5, nudge_y = 0.005) -> saveme
png("SMG8_comparison_PSI_correlations_both.png", height=3, width=3, units="in", res=1200)
saveme
dev.off()
```