---
title: "siUPF1 DTE ECDF plots"
output:
  html_document:
    df_print: paged
---

The purpose of this script is to look at how population classes of 3'UTR spliced transcripts react to UPF1 knockdown

```{r}
library(DESeq2)
library(tximport)
library(dplyr)
library(stringr)
```

Load sample names

```{r}
tx2gene<-read.delim(file='../../../../../HCT-NMD_KD/salmonquant/transcript2geneMap.tsv', 
                    header = TRUE, sep = '\t', dec = '.')

quant_dir = "../../../../../HCT-NMD_KD/salmonquant/quantification.dir"
all_files = list.files(quant_dir, pattern="*.sf")
all_files = file.path(quant_dir,  all_files)

sample_names <- sapply(all_files, function(x) substr(x, 0, nchar(x)-3))
sample_names <- sapply(sample_names, function(x) substr(x, 32, nchar(x)))
names(all_files) <- sample_names
```

Generate sample table, comparing siUPF1 to control

```{r}
sample_table = data.frame(files=all_files, names=sample_names)
sample_table = sample_table %>% mutate(condition=ifelse(stringr::str_detect(names, "dsRed"), "siControl", "siUPF1"))
sample_table$files = as.character(sample_table$files)
```

Tximport transcript level

```{r}
txi <- tximport(sample_table$files, type="salmon", tx2gene=tx2gene, txOut=T)
```

Generate deseq dataset to compare the conditions

```{r}
dds <- DESeqDataSetFromTximport(txi, sample_table, ~condition)
```

Run deseq

```{r}
dds_result = DESeq(dds)
```

Filter for rows where there is at least 1 count in all samples for at least 1 condition

```{r}
nrow(dds_result)
keep = rowSums(counts(dds_result)[,1:8]>1)>=4|rowSums(counts(dds_result)[,9:16]>1)>=4
sum(keep)
dds_result = dds_result[keep,]
nrow(dds_result)
```

Deseq2 result

```{r}
dds_result = results(dds_result)
```

MA plot

```{r}
plotMA(dds_result)
```

Filter for utrons

```{r}
library(data.table)
utron_ids = fread("../../../all_TCGA/utron_beds.dir/all_tissues_agg-agg-agg.all_utrons.bed") 
utron_ids = utron_ids %>% dplyr::select(V4) %>% distinct() %>% unlist() %>% unname()
result_df = as.data.frame(dds_result)
result_df_has_utron = setDT(result_df, keep.rownames=T) %>%
  filter(rn %in% utron_ids)
row.names(result_df_has_utron) = result_df_has_utron$rn
result_df_has_utron = left_join(result_df_has_utron, tx2gene, by=c("rn"="transcript_id"))
```

```{r}
library(AnnotationDbi)
library(org.Hs.eg.db)
result_df_has_utron$gene_id = as.character(result_df_has_utron$gene_id)
result_df_has_utron$symbol = mapIds(org.Hs.eg.db,
                           keys=result_df_has_utron$gene_id,
                           column="SYMBOL",
                           keytype="ENSEMBL",
                           multiVals="first")
```

Lets calc stop distance for each of the utrons

```{r}
stop_codons = fread("../../../all_TCGA/expression.dir/gtf_stop_codons.txt", col.names=c("gene_id", "transcript_id", "symbol", "strand", "stop_codon", "chromosome", "start", "end"))
splice_sites = fread("../../../all_TCGA/expression.dir/all_splice_sites.txt")
splice_site_distances <- splice_sites %>%
  inner_join(result_df_has_utron, by=c("transcript_id"="rn")) %>%
  inner_join(dplyr::select(stop_codons, transcript_id, start, end)) %>% 
  mutate(stop_distance = ifelse(strand=="+",
                            splice_site_start - end,
                            start - splice_site_end))%>%
  dplyr::select(transcript_id,
         splice_site_start,
         splice_site_end,
         stop_distance) %>%
  distinct()

#mutate to give bool for greater than 55
splice_site_distances = splice_site_distances %>% mutate(`>55nt`=stop_distance>55)
```

### ECDF plots

```{r}
library(ggplot2)
non_cds <- read.delim("../../../../../Cell_lines/CCLE/new_utrons_classification/utron_beds.dir/agg-agg-agg.no_cds.ids.gz", header=F)
plot_data = result_df_has_utron %>% inner_join(splice_site_distances, by=c("rn"="transcript_id")) %>%
  mutate(non_CDS = rn %in% non_cds$V1)

cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Lets look at cds (PTC) vs non cds (nonPTC)

```{r, fig.height=2, fig.width=2}
over55 = plot_data %>% filter(`>55nt`) %>% dplyr::select(log2FoldChange) %>% unlist()
under55 = plot_data %>% filter(`>55nt`==F) %>% dplyr::select(log2FoldChange) %>% unlist()
test = ks.test(over55, under55)
plot_data %>% mutate(class = ifelse(`>55nt`, ">55nt", "<55nt")) %>%
ggplot() +
  aes(x=log2FoldChange, col=class) + 
  stat_ecdf() + 
  coord_cartesian(xlim=c(-5,5)) + 
  scale_color_manual(values=cbbPalette) + 
  theme_bw(base_size = 8) + 
  theme(legend.position = "top", 
        legend.text = element_text(size=8),
        legend.title = element_blank(),
        axis.text = element_text(size=8)) + ylab("") + xlab("log2FoldChange(siUPF1)")+ 
  annotate("text", label=paste0("D=",round(test$statistic,3)), x=-5, y=0.95, hjust=0) + 
  annotate("text", label=paste0("p=",round(test$p.value,5)), x=-5, y=0.8, hjust=0) -> saveme
png("over_vs_under_55.png", width=2, height=2, units="in", res=1200)
saveme
dev.off()
```

Now lets look at predicted nmd sensitive vs non-sensitive for nonPTC 3UIs, i.e. is it a case that <55 go down, >55nt go up, and therefore it balances in the above plot?

```{r, fig.height=2, fig.width=2}
library(RSQLite)
biotypes <- dbGetQuery(dbConnect(SQLite(), "../../General/annotations/hg38_noalt_ensembl85/csvdb"),
                          "SELECT DISTINCT transcript_id, transcript_biotype FROM transcript_info")

all_utrons <- read.delim("../../../../../Cell_lines/CCLE/new_utrons_classification/utron_beds.dir/agg-agg-agg.all_utrons.ids.gz", header=F)
novel_utrons <- read.delim("../../../../../Cell_lines/CCLE/new_utrons_classification/utron_beds.dir/agg-agg-agg.novel_utrons.ids.gz", header=F)
partnered_utrons <- read.delim("../../../../../Cell_lines/CCLE/new_utrons_classification/utron_beds.dir/agg-agg-agg.partnered_utrons.ids.gz", header=F)
non_cds <- read.delim("../../../../../Cell_lines/CCLE/new_utrons_classification/utron_beds.dir/agg-agg-agg.no_cds.ids.gz", header=F)

utron_annotation <- data.frame(  transcript_id = result_df$rn) %>%
  mutate(utron = transcript_id %in% all_utrons$V1,
         novel_utron = transcript_id %in% novel_utrons$V1,
         partnered_utron = transcript_id %in% partnered_utrons$V1,
         non_cds = transcript_id %in% non_cds$V1)

plot_data = result_df %>% inner_join(utron_annotation, by=c("rn"="transcript_id")) %>%
  left_join(biotypes, by=c("rn"="transcript_id")) %>%
  mutate(transcript_biotype = ifelse(is.na(transcript_biotype), "Novel", transcript_biotype)) %>%
  mutate(category=ifelse(transcript_biotype=="protein_coding" & ! utron, "Coding",
                         ifelse((transcript_biotype == "nonsense_mediated_decay" | utron) & !non_cds, "NMD",
                         ifelse(non_cds, "nPTC", NA))))

test_nonPTC = plot_data %>% filter(non_cds) %>% dplyr::select(log2FoldChange) %>% unlist()
test_pPTC = plot_data %>% filter(non_cds==F) %>% dplyr::select(log2FoldChange) %>% unlist()
test = ks.test(test_nonPTC, test_pPTC)

plot_data %>% filter(category!="NA") %>%
  ggplot() +
  aes(x=log2FoldChange, col=category) + 
  stat_ecdf() + 
  coord_cartesian(xlim=c(-5,5)) + 
  scale_color_manual(values=cbbPalette)+ 
  theme_bw(base_size = 8) + 
  theme(legend.position = "top", 
        legend.text = element_text(size=8),
        legend.title = element_blank(),
        axis.text = element_text(size=8),
        legend.key.width = unit(3,"mm")) + ylab("") + xlab("log2FoldChange(siUPF1)")+ 
  annotate("text", label=paste0("D=",round(test$statistic,3)), x=-5, y=0.95, hjust=0) + 
  annotate("text", label=paste0("p=",round(test$p.value, 12)), x=-5, y=0.8, hjust=0) -> saveme
png("ptc_vs_nonPTC.png", height=2, width=2, units="in", res=1200)
saveme
dev.off()
```